---
knit: pagedown::chrome_print
output:
  pagedown::html_paged:
    self_contained: TRUE
    number_sections: FALSE
    fig_caption: TRUE
    css: ["output_styles.css"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, message = FALSE, warning=FALSE, include = FALSE}
library(tidyverse) #loads multiple packages for BP data management, viz, and more. Should always include
library(janitor) #for cleaning data
library(ggplot2) # package to make visuals
library(rlang) # package to create functions that call column names
library(knitr) # package to style tables
library(scales) # package to clean up plot labels
```

```{r load data, include = FALSE}
load("../Output/last_seven_days_map_data.rdata")
load("../Output/moving_average_data.rdata")
```

```{r set up trend chart function}
trend_chart <- function(data, moving_average_variable, color_values,
                        y_label_format, legend_position){
  data %>%
  ggplot(aes(x = date, y = {{ moving_average_variable }}, color = state)) +
    # reverse order of the states so Colorado is plotted last
    aes(group=rev(state)) + 
    # assign the colors to each line
    scale_colour_manual(values = color_values) +
    # set thickness of the line
    geom_line(size=0.5) + 
    # adjust the x-axis to mark the first of each month
    scale_x_date(
      date_breaks = "1 month",
      breaks = as.Date(c("2020-03-01", "2020-10-01")),
      minor_breaks = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01",
                               "2020-07-01", "2020-08-01", "2020-09-01",
                               "2020-10-01")),
      date_labels = "%B") +
    # adjust the y-axis label
    scale_y_continuous(
      label = y_label_format
    ) +
    # clean up the layout of the chart
    theme_minimal() +
    theme(axis.title = element_blank(),
          legend.position = legend_position,
          legend.title = element_blank())
}
```

# Final Visuals
<center>By: Mayra Smith-Coronado</center>
<center>`r format(Sys.Date(), '%d %B, %Y')`</center>

## Exercise 1

Use data from the CDC available for download from [here](https://covid.cdc.gov/covid-data-tracker/#cases) to create a map showing US COVID-19 case incidence per 100,000 in the last 7 days. You'll also need to find and use state-level population estimates. Please only include the 48 continental US states, Alaska, and Hawaii in your final figure.

```{r cars}
last_seven_days_map_data  %>%
  ggplot(aes(fill = rate_discrete)) +
  # add state borders 
  geom_sf(
    color = "gray50") +
  # crop the map to be zoomed in on locations
  coord_sf(
    xlim = c(-185, -60),
    ylim = c(15, 75),
    expand = FALSE) +
  # update color & reverse legend
  scale_fill_manual(
    values = c("#EDEDEF", "#E2DDDA", "#c7cfd1", "#9dadb0", "#789093", "#33565C"),
    guide = guide_legend(reverse = TRUE, title = "Cases per 100,000")) + 
  # clean up map image
  theme_minimal() +
  theme(
    axis.line = element_blank(), # remove axis line 
    axis.text = element_blank(), # remove axis text
    panel.grid = element_blank()
  )
```

\newpage

## Exercise 2

Use state-level historical data from the COVID Tracking Project to visualize

- The 7-day moving average of new cases reported per 100,000
- The 7-day moving average of daily PCR tests conducted per 100,000 people
- The 7-day moving average of percent positivity (person-level positive PCR tests divided by the total number of people who received a PCR test each day)

---

Since June, <span style = "color: #1F4045;font-weight: bold;">Colorado's</span> 7-day moving average of new cases reported per 100,000 people is estimated to be lower than neighboring states, including Texas. 

```{r new cases chart, fig.height = 2.65, fig.width = 6, fig.align = "center", echo=FALSE, message = FALSE, warning=FALSE}
trend_chart(
  data = moving_average,
  moving_average_variable = new_case_moving_ave,
  color_values = c("#1F4045", rep("#EBCDC3", times = 8)),
  y_label_format = comma,
  legend_position = "right"
)
```

---

The number of daily PCR tests conducted per 100,000 people continues to grow in <span style = "color: #1F4045;font-weight: bold;">Colorado</span>.

```{r PCR chart, fig.height = 2.65, fig.width = 6, fig.align = "center", echo=FALSE, message = FALSE, warning=FALSE}
co_moving_average = moving_average %>%
  filter(state == "CO")

trend_chart(
  data = co_moving_average,
  moving_average_variable = pcr_test_moving_ave,
  color_values = c("#1F4045"),
  y_label_format = comma,
  legend_position = "none"
)
```

\newpage

The percent positivity continues to decrease in <span style = "color: #1F4045;font-weight: bold;">Colorado</span>.

```{r percent positivity, fig.height = 2.65, fig.width = 6, fig.align = "center", echo=FALSE, message = FALSE, warning=FALSE}
trend_chart(
  data = co_moving_average,
  moving_average_variable = percent_positivity_moving_ave,
  color_values = c("#1F4045"),
  y_label_format = label_percent(),
  legend_position = "none")
```

